---
title: "Untitled"
format: html
editor: visual
---

## A structural Merton jump-diffusion framework for survival analysis: Modeling biological solvency and distance-to-death (DtD) in tuberculosis

```{r}
############################################################
## 0. PACKAGES
############################################################
library(dplyr)
library(survival)
library(ggplot2)

library(knitr)
library(kableExtra)

library(MASS)
library(numDeriv)

```

## 1. Importing data

```{r}
## IMPORT DATA
path <- "https://raw.githubusercontent.com/pefura/structural-jump-diffusion/refs/heads/main/synthetic_tb_data.csv"
df_clean2 <- read.csv(path)
head(df_clean2)
```

## 2. Data preparation, Parameters, Merton and Extended Cox

```{r}
# Random fixed
set.seed(123)

############################################################
## 1. DATA PREPARATION & TRAIN / TEST SPLIT
############################################################

# Define categorical and numeric variables
cat_var <- c("issue_traitement", "lieusuivi",
             "TB_class", "HIV_status")
num_var <- c("surv_time_day", "agecode", "BMI")

# Convert variables
df_clean2[cat_var] <- lapply(df_clean2[cat_var], factor)
df_clean2[num_var] <- lapply(df_clean2[num_var], as.numeric)

# Define survival variables
df_clean2 <- df_clean2 %>%
  mutate(
    Duration = surv_time_day,
    Status   = ifelse(issue_traitement == "death", 1, 0)
  )

# Train / test split
n <- nrow(df_clean2)
idx <- sample(seq_len(n), size = 0.7 * n)
train <- df_clean2[idx, ]
test  <- df_clean2[-idx, ]

############################################################
## 2. RECODING CLINICAL VARIABLES – TB REFERENCE EXPLICIT
############################################################

# Ensure "smear positive pulmonary" is reference
train <- train %>%
  mutate(
    VIH      = factor(HIV_status, levels = c("negative", "positive")),
    Hospit   = factor(lieusuivi, levels = c("outpatient", "hospitalized")),
    TB_class = factor(
      TB_class,
      levels = c("smear positive pulmonary",  # REFERENCE
                 "extrapulmonary",
                 "smear negative pulmonary")
    )
  )

test <- test %>%
  mutate(
    VIH      = factor(HIV_status, levels = c("negative", "positive")),
    Hospit   = factor(lieusuivi, levels = c("outpatient", "hospitalized")),
    TB_class = factor(
      TB_class,
      levels = c("smear positive pulmonary",  # REFERENCE
                 "extrapulmonary",
                 "smear negative pulmonary")
    )
  )

############################################################
## 3. MERTON NEGATIVE LOG-LIKELIHOOD (RAW AGE & BMI)
############################################################

merton_nll <- function(params, data){

  mu_base     <- params[1]
  beta_age    <- params[2]
  sigma_base  <- params[3]
  lambda_base <- params[4]
  K           <- params[5]
  delta       <- params[6]

  n_TB <- length(levels(data$TB_class)) - 1
  gamma_TB <- params[7:(6 + n_TB)]

  coeff_VIH  <- params[7 + n_TB]
  coeff_Hosp <- params[8 + n_TB]

  T_hor <- 240 / 365

  # Personalized drift (RAW AGE)
  mu_i <- mu_base + beta_age * data$agecode
  for (l in seq_len(n_TB)) {
    mu_i <- mu_i + gamma_TB[l] *
      (data$TB_class == levels(data$TB_class)[l + 1])
  }

  # Volatility and jump intensity
  sigma_i  <- sigma_base * ifelse(data$VIH == "positive", coeff_VIH, 1)
  lambda_i <- lambda_base * ifelse(data$Hospit == "hospitalized", coeff_Hosp, 1)

  # Aggregated dynamics
  sigma_tot <- sqrt(sigma_i^2 + lambda_i * delta^2)
  mu_adj    <- mu_i - lambda_i * (exp(delta^2 / 2) - 1)

  # DTD score (RAW BMI)
  DTD <- (log(data$BMI / K) +
          (mu_adj - 0.5 * sigma_tot^2) * T_hor) /
         (sigma_tot * sqrt(T_hor))

  prob_surv <- pnorm(DTD)
  prob_surv <- pmax(pmin(prob_surv, 0.999), 0.001)

  nll <- -sum(data$Status * log(1 - prob_surv) +
              (1 - data$Status) * log(prob_surv))

  return(nll)
}

############################################################
## 4. PARAMETER OPTIMIZATION
############################################################

n_TB <- length(levels(train$TB_class)) - 1

init_params <- c(
  mu_base = 0.05,
  beta_age = -0.01,
  sigma_base = 0.25,
  lambda_base = 0.05,
  K = 18,
  delta = 0.15,
  rep(0.02, n_TB),
  coeff_VIH = 1.5,
  coeff_Hosp = 3
)

fit <- optim(
  par = init_params,
  fn = merton_nll,
  data = train,
  method = "L-BFGS-B",
  lower = c(-1, -0.2, 0.05, 0.001, 15, 0.05, rep(-1, n_TB), 0.5, 1),
  upper = c(10,  0.1,  2,    2,    30,  2,    rep(1, n_TB), 5,   10),
  control = list(maxit = 10000)
)

if(fit$convergence == 0){
  cat("Successful convergence.\n")
} else {
  stop("Optimization did not converge")
}

opt_params <- fit$par

# Annotate gamma_TB coefficients
gamma_TB <- opt_params[7:(6 + n_TB)]
tb_levels <- levels(train$TB_class)[-1]  # exclude reference
names(gamma_TB) <- tb_levels

cat("\nParamètres Merton optimisés :\n")
print(opt_params)
cat("\nGamma_TB coefficients (relative to 'smear positive pulmonary'):\n")
print(gamma_TB)

############################################################
## 5. DTD COMPUTATION
############################################################

calc_DTD <- function(df, params){

  mu_base     <- params[1]
  beta_age    <- params[2]
  sigma_base  <- params[3]
  lambda_base <- params[4]
  K           <- params[5]
  delta       <- params[6]

  n_TB <- length(levels(df$TB_class)) - 1
  gamma_TB <- params[7:(6 + n_TB)]
  coeff_VIH <- params[7 + n_TB]
  coeff_Hosp <- params[8 + n_TB]

  T_hor <- 240 / 365

  mu_i <- mu_base + beta_age * df$agecode
  for (l in seq_len(n_TB)) {
    mu_i <- mu_i + gamma_TB[l] *
      (df$TB_class == levels(df$TB_class)[l + 1])
  }

  sigma_i  <- sigma_base * ifelse(df$VIH == "positive", coeff_VIH, 1)
  lambda_i <- lambda_base * ifelse(df$Hospit == "hospitalized", coeff_Hosp, 1)

  sigma_tot <- sqrt(sigma_i^2 + lambda_i * delta^2)
  mu_adj    <- mu_i - lambda_i * (exp(delta^2 / 2) - 1)

  DTD <- (log(df$BMI / K) +
          (mu_adj - 0.5 * sigma_tot^2) * T_hor) /
         (sigma_tot * sqrt(T_hor))

  list(
    DTD_score = DTD,
    prob_surv = pnorm(DTD)
  )
}

train_res <- calc_DTD(train, opt_params)
test_res  <- calc_DTD(test, opt_params)

train$DTD_score <- train_res$DTD_score
train$prob_surv <- train_res$prob_surv

test$DTD_score <- test_res$DTD_score
test$prob_surv <- test_res$prob_surv

############################################################
## 6. EXTENDED COX MODEL
############################################################

train <- train %>%
  mutate(
    VIH_pos = as.numeric(VIH == "positive"),
    Hospit_hosp = as.numeric(Hospit == "hospitalized")
  )

test <- test %>%
  mutate(
    VIH_pos = as.numeric(VIH == "positive"),
    Hospit_hosp = as.numeric(Hospit == "hospitalized")
  )

cox_ext <- coxph(
  Surv(Duration, Status) ~ agecode + BMI + VIH_pos +
    Hospit_hosp + TB_class +
    tt(VIH_pos) + tt(Hospit_hosp),
  data = train,
  tt = function(x, t, ...) x * log(t + 1)
)

############################################################
## 7. C-INDEX WITH BOOTSTRAP
############################################################

calc_c <- function(s, score){
  c <- concordance(s ~ score)$concordance
  if(c < 0.5) c <- 1 - c
  c
}

B <- 1000
set.seed(123)

c_m <- c_c <- delta_c <- numeric(B)
surv_test <- Surv(test$Duration, test$Status)

for(b in 1:B){
  idx_b <- sample(seq_len(nrow(test)), replace = TRUE)
  surv_b <- Surv(test$Duration[idx_b], test$Status[idx_b])

  c_m[b] <- calc_c(surv_b, -test$DTD_score[idx_b])
  c_c[b] <- calc_c(
    surv_b,
    predict(cox_ext, newdata = test[idx_b, ], type = "lp")
  )

  delta_c[b] <- c_m[b] - c_c[b]
}

############################################################
## 8. RESULTS SUMMARY
############################################################

cat("\nC-index Merton (95% CI):",
    round(mean(c_m), 3),
    quantile(c_m, c(0.025, 0.975)), "\n")

cat("C-index Cox (95% CI):",
    round(mean(c_c), 3),
    quantile(c_c, c(0.025, 0.975)), "\n")

cat("ΔC (Merton − Cox):",
    round(mean(delta_c), 3),
    quantile(delta_c, c(0.025, 0.975)), "\n")

cat("Bootstrap p-value:",
    min(mean(delta_c <= 0) * 2, 1), "\n")

############################################################
## 9. DTD SCORE SUMMARIES
############################################################

cat("\nTrain DTD score summary:\n")
print(summary(train$DTD_score))

cat("\nTest DTD score summary:\n")
print(summary(test$DTD_score))

cat("\nTrain probability of survival summary:\n")
print(summary(train$prob_surv))

cat("\nTest probability of survival summary:\n")
print(summary(test$prob_surv))
```

## 3. Merton parameters

```{r}
############################################################
## AUTOMATIC MERTON PARAMETER TABLE
############################################################

# Number of TB coefficients
n_TB <- length(levels(train$TB_class)) - 1
tb_levels <- levels(train$TB_class)[-1]  # exclude reference

# Build parameter names (mathematical + readable)
param_names <- c(
  "$\\mu_0$ (baseline drift)",
  "$\\beta_{age}$ (age effect)",
  "$\\sigma_0$ (baseline volatility)",
  "$\\lambda_0$ (baseline jump intensity)",
  "$K$ (nutritional threshold)",
  "$\\delta$ (jump size)",
  paste0("$\\gamma_{TB}$: ", tb_levels),
  "$\\theta_{HIV}$ (HIV multiplier)",
  "$\\theta_{Hosp}$ (hospitalization multiplier)"
)

# Create table
merton_table <- data.frame(
  Parameter = param_names,
  Estimate  = round(opt_params, 4)
)

# Add reference note
attr(merton_table, "note") <- 
  "TB reference category: smear positive pulmonary"

kable(
  merton_table,
  format = "html",
  caption = "Optimized Merton model parameters",
  escape = FALSE
) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) %>%
  footnote(
    general = attr(merton_table, "note")
  )

```

## **4. Optimized structural parameters and robust statistical inference of the Merton jump-diffusion model**

```{r}
################################################################################
## 10. ROBUST STATISTICAL INFERENCE (FOR 10-PARAMETER MODEL)
################################################################################

# 1. Calculate a precise Hessian matrix using Richardson extrapolation
# This is more reliable for p-values than the default optim output
hess_precise <- numDeriv::hessian(func = merton_nll, x = fit$par, data = train)

# 2. Robust matrix inversion (Moore-Penrose Pseudo-inverse)
# This handles numerical instability often found in SDE models
var_cov <- ginv(hess_precise)

# 3. Calculation of Standard Errors
# We use abs() to ensure positive values before the square root
se <- sqrt(abs(diag(var_cov)))

# 4. Statistical Inference (Wald tests)
# Calculate z-scores and two-sided p-values assuming asymptotic normality
z_val <- fit$par / se
p_val <- 2 * (1 - pnorm(abs(z_val)))

# 5. Define Parameter Names (Matching your merton_nll function mapping)
# 1:mu_base, 2:age, 3:sigma, 4:lambda, 5:K, 6:delta, 7:EPTB, 8:SNPTB, 9:VIH, 10:Hosp
param_names <- c("mu_base", "beta_age", "sigma_base", "lambda_base", "K", "delta", 
                 "gamma_EPTB", "gamma_SNPTB", "coeff_VIH", "coeff_Hosp")

# 6. Construct Results Table
results_merton <- data.frame(
  Parameter = param_names,
  Estimate  = round(fit$par, 4),
  Std_Error = round(se, 4),
  z_value   = round(z_val, 2),
  p_value   = format.pval(p_val, digits = 3, eps = 0.001)
)

# 7. Print the final summary table
cat("\n--- Optimized Merton Parameters (10-Parameter Model) ---\n")
print(results_merton)
```

## 5. Full model vs reduced model

```{r}
################################################################################
## MODEL COMPARISON: FULL MODEL (JUMP-DIFFUSION) VS. REDUCED MODEL (PURE DIFFUSION)
################################################################################

# 1. Define the Reduced Model (Pure Diffusion - NO JUMPS)
# This model assumes health trajectories are continuous without acute shocks.
merton_nll_reduced <- function(params, data){
  mu_base     <- params[1]
  beta_age    <- params[2]
  sigma_base  <- params[3]
  K           <- params[4]
  
  n_TB        <- length(levels(data$TB_class)) - 1
  gamma_TB    <- params[5:(4 + n_TB)]
  coeff_VIH   <- params[5 + n_TB]

  T_hor <- 240 / 365

  # Personalized Drift (mu_i)
  mu_i <- mu_base + beta_age * data$agecode
  for (l in seq_len(n_TB)) {
    mu_i <- mu_i + gamma_TB[l] * (data$TB_class == levels(data$TB_class)[l + 1])
  }

  # Pure Volatility (No Jump components: Lambda * Delta excluded)
  sigma_i <- sigma_base * ifelse(data$VIH == "positive", coeff_VIH, 1)

  # Simplified Distance-to-Death (DtD) score (Standard Diffusion/Black-Scholes type)
  DTD <- (log(data$BMI / K) + (mu_i - 0.5 * sigma_i^2) * T_hor) / (sigma_i * sqrt(T_hor))

  # Calculate Survival Probability
  prob_surv <- pnorm(DTD)
  prob_surv <- pmax(pmin(prob_surv, 0.999), 0.001)

  # Calculate Negative Log-Likelihood
  nll <- -sum(data$Status * log(1 - prob_surv) + (1 - data$Status) * log(prob_surv))
  return(nll)
}

# 2. Optimization of the Reduced Model
init_params_red <- c(
  mu_base = 0.05, beta_age = -0.01, sigma_base = 0.25, 
  K = 18, rep(0.02, n_TB), coeff_VIH = 1.5
)

fit_red <- optim(
  par = init_params_red,
  fn = merton_nll_reduced,
  data = train,
  method = "L-BFGS-B",
  lower = c(-1, -0.2, 0.05, 15, rep(-1, n_TB), 0.5),
  upper = c(10,  0.1,  2,   30, rep(1, n_TB), 5),
  control = list(maxit = 10000)
)

# 3. Calculate Model Selection Criteria and Likelihood Ratio Test (LRT)
k_full   <- length(fit$par)   # Number of parameters in Full Model
k_red    <- length(fit_red$par) # Number of parameters in Reduced Model
nll_full <- fit$value         # NLL of Full Model (previously calculated)
nll_red  <- fit_red$value     # NLL of Reduced Model

# Information Criteria
aic_red  <- 2 * k_red + 2 * nll_red
bic_red  <- log(nrow(train)) * k_red + 2 * nll_red

# Likelihood Ratio Test Statistic
# Calculated as 2 * (difference in log-likelihoods)
lrt_stat <- 2 * (nll_red - nll_full) 
df_diff  <- k_full - k_red    # Degrees of freedom difference
p_lrt    <- pchisq(lrt_stat, df = df_diff, lower.tail = FALSE)

# 4. Display Comparative Results Summary
cat("\n================================================")
cat("\n           MODEL COMPARISON SUMMARY")
cat("\n================================================")
cat("\nFULL MODEL (Jump-Diffusion):")
cat("\n  - AIC :", round(4596.92, 2)) 
cat("\n  - NLL :", round(nll_full, 2))
cat("\n")
cat("\nREDUCED MODEL (Pure Diffusion):")
cat("\n  - AIC :", round(aic_red, 2))
cat("\n  - NLL :", round(nll_red, 2))
cat("\n------------------------------------------------")
cat("\nLikelihood Ratio Test (LRT):")
cat("\n  - Chi-squared stat :", round(lrt_stat, 4))
cat("\n  - Degrees of freedom :", df_diff)
cat("\n  - p-value :", format.pval(p_lrt, eps = 0.001))
cat("\n------------------------------------------------")

if(p_lrt < 0.05) {
  cat("\nDECISION: The FULL MODEL is statistically superior.")
  cat("\nJustification: Jump components contribute significant explanatory power.")
} else {
  cat("\nDECISION: The REDUCED MODEL is preferred (Parsimony).")
  cat("\nJustification: Jumps do not significantly improve model fit.")
}
cat("\n================================================\n")
```

## 6. Graphics

```{r}
# ==============================================================================
# 3. RESULTS & VISUALIZATION (UPDATED: START AT 0 & BOLD AXES)
# ==============================================================================

# Determine a reasonable upper limit for the plot (e.g., 0.25 or 0.30)
max_val <- max(c(calib_merton$mean_pred, calib_merton$surv_obs, 
                 calib_cox$mean_pred, calib_cox$surv_obs), na.rm = TRUE) * 1.1

p <- ggplot() +
  # Reference line (Perfect Calibration)
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "black", alpha = 0.6) +
  
  # Merton (Blue)
  geom_line(data = calib_merton, aes(x = mean_pred, y = surv_obs, color = "Merton jump-diffusion"), size = 1.2) +
  geom_point(data = calib_merton, aes(x = mean_pred, y = surv_obs, color = "Merton jump-diffusion"), size = 3) +
  
  # Cox (Red)
  geom_line(data = calib_cox, aes(x = mean_pred, y = surv_obs, color = "Extended Cox"), size = 1.2) +
  geom_point(data = calib_cox, aes(x = mean_pred, y = surv_obs, color = "Extended Cox"), size = 3) +
  
  # Aesthetics
  scale_color_manual(values = c("Merton jump-diffusion" = "blue", "Extended Cox" = "red")) +
  
  # FORCING AXES TO START AT 0
  # expand = c(0,0) removes the padding R usually adds around the 0 point
  scale_x_continuous(limits = c(0, max_val), expand = c(0, 0)) +
  scale_y_continuous(limits = c(0, max_val), expand = c(0, 0)) +
  
  # Ensure the plot is a perfect square so the 45-degree line is accurate
  coord_equal() +
  
  labs(
    #title = paste("Calibration Plot at", T_target, "Days"),
    #subtitle = paste0("GND Chi2: Merton=", round(GND_chi_merton, 1), " vs Cox=", round(GND_chi_cox, 1)),
    x = "Predicted probability of death",
    y = "Observed probability of death",
    color = "Model"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    # BOLDING THE MAIN AXES TITLES
    axis.title.x = element_text(size = 12, margin = margin(t = 10)),
    axis.title.y = element_text(size = 12, margin = margin(r = 10)),
    # BOLDING THE NUMBERS (Axis Text)
    axis.text.x = element_text(face = "bold", size = 10),
    axis.text.y = element_text(face = "bold", size = 10),
    # BOLD SOLID AXIS LINES
    axis.line = element_line(color = "black", size = 1.1),
    # Centering Titles
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    # Cleaning background
    panel.grid.minor = element_blank()
  )

# Save the plot
ggsave("/Users/pefura-yone/Documents/FMSB/Articles/Stochastic_modeling_TB/calibration_plot_final_legend.png", 
       plot = p, width = 9, height = 7, dpi = 300)

# Display
print(p)
```

```{r}
################################################################################
## FIGURE 2: MULTIPANEL CLINICAL VALIDATION (FIXED LABELS & HIGH CONTRAST)
################################################################################

# 1. Data Preparation
test <- test %>%
  mutate(
    Outcome = factor(Status, labels = c("Survivor", "Dead")),
    DTD_quartile = ntile(DTD_score, 4),
    DTD_quartile = factor(DTD_quartile, 
                          labels = c("Q1: High Risk", "Q2", "Q3", "Q4: Low Risk")),
    HIV_Status = factor(VIH, labels = c("HIV Negative", "HIV Positive"))
  )

# High-contrast palette: Red, Orange, Blue, Green
km_cols <- c("#d73027", "#f46d43", "#4575b4", "#1a9850")

# --- Panel A: Distribution Separation ---
p_dist <- ggplot(test, aes(x = Outcome, y = DTD_score, fill = Outcome)) +
  geom_violin(alpha = 0.3, color = NA) +
  geom_boxplot(width = 0.2, outlier.shape = 16, outlier.alpha = 0.2, color = "black", size = 0.6) +
  scale_fill_manual(values = c("Survivor" = "#4575b4", "Dead" = "#d73027")) +
  labs(title = "A. DtD separation by outcome", y = "Distance-to-Death (DtD)", x = "") +
  theme_bw() + theme(legend.position = "none", axis.title.y = element_text(face="bold"))

# --- Panel B: Model Granularity (BMI vs DtD) ---
p_scatter <- ggplot(test, aes(x = BMI, y = DTD_score, color = HIV_Status)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(method = "lm", formula = y ~ x, color = "black", linetype = "dashed", size = 0.7, se = FALSE) +
  scale_color_manual(values = c("HIV Negative" = "#1a9850", "HIV Positive" = "#d73027")) +
  labs(title = "B. BMI vs. DtD granularity", x = "Admission BMI (kg/m²)", y = "DtD score", color = "HIV Status") +
  theme_bw() + theme(legend.position = "bottom", axis.title = element_text(face="bold"))

# --- Panel C: Kaplan-Meier (FIXED Y-AXIS LABEL) ---
fit_q <- survfit(Surv(Duration, Status) ~ DTD_quartile, data = test)
p_km_obj <- ggsurvplot(fit_q, data = test, 
                       fun = "pct", palette = km_cols, 
                       ylim = c(70, 100), # Plot starts at 70%
                       title = "C. Survival by DtD quartile",
                       xlab = "Time (days)", ylab = "Survival (%)",
                       legend.title = "Risk stratum",
                       legend.labs = c("Q1", "Q2", "Q3", "Q4"),
                       ggtheme = theme_bw())

p_km <- p_km_obj$plot + 
  # Placement at y = 75 ensures it shows up within the 70-100 range
  annotate("text", x = 10, y = 75, label = "Log-rank p < 0.0001", 
           hjust = 0, size = 3.5, fontface = "bold.italic") +
  theme(axis.title = element_text(face="bold"))

# --- Panel D: Mortality Gradient ---
mort_data <- data.frame(Quartile = c("Q1", "Q2", "Q3", "Q4"), Rate = c(16.7, 7.5, 1.8, 1.6))
p_grad <- ggplot(mort_data, aes(x = Quartile, y = Rate, fill = Quartile)) +
  geom_bar(stat = "identity", width = 0.6, color = "black", size = 0.3) +
  scale_fill_manual(values = km_cols) + 
  geom_text(aes(label = paste0(Rate, "%")), vjust = -0.5, fontface = "bold", size = 3.5) +
  labs(title = "D. Mortality gradient", x = "DtD quartile", y = "Mortality rate (%)") +
  theme_bw() + theme(legend.position = "none", axis.title = element_text(face="bold")) +
  ylim(0, 22)

# --- Final Assembly ---
final_fig2 <- (p_dist | p_scatter) / (p_km | p_grad) + plot_layout(heights = c(1, 1))

ggsave("Figure2_DtD_Validation_Final.png", plot = final_fig2, width = 10, height = 9, dpi = 300)
print(final_fig2)
```
